\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-27 chapters/gaze_independence.tex
Bleed

The interface can be operated in different \ac{vsa} conditions, as illustrated in
\cref{fig:gaze/vsa}.

\begin{figure}
  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_overt.pdf}
    \caption[Overt \ac{vsa}]{%
      \emph{Overt} \ac{vsa}.
      Gaze and \ac{vsa} coincide on a target.
    }
    \label{fig:gaze/vsa/overt}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_covert.pdf}
    \caption[Covert \ac{vsa}]{%
      \emph{Covert} \ac{vsa}.
      Gaze rests on the center of the screen, while \ac{vsa} is directed towards a target.
    }
    \label{fig:gaze/vsa/covert}
  \end{subfigure}

  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.pdf}
    \caption[Split \ac{vsa}]{%
      \emph{Split} \ac{vsa}.
      \Ac{vsa} is directed towards a target, while the gaze rests on another.
    }
    \label{fig:gaze/vsa/split}
  \end{subfigure}\hfill%
   \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.pdf}
    \caption[Free \ac{vsa}]{%
      \emph{Free} \ac{vsa}.
      The user is free to direct their gaze as they deem most comfortable.
      All previous \ac{vsa} conditions are possible.
    }
    \label{fig:gaze/vsa/free}
  \end{subfigure}\hfill%
  \caption[\Ac{vsa} conditions]{%
    \Acf{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm
    interface. Full lines indicated gaze fixation, dashed lines \ac{vsa}.
  }
  \label{fig:gaze/vsa}
\end{figure}

In the overt case, users gazed at the cued
target they were also mentally attending; in the covert case, users
gazed at the center of the screen while mentally attending to the cued target.
We introduced a \ac{vsa} condition that is understudied in the context of
gaze-independent \ac{bci} development: split \ac{vsa}.
In split \ac{vsa}, the user mentally focuses on one cued target while gazing at
another (the distractor).
This last option has been scarcely studied but completes
the options to dissociate gaze and visuospatial attention, allowing us to
investigate the effect of (the lack of) gaze control on \ac{bci} performance.
This dataset and the performance of our proposed \ac{wcble} decoder have also been
published by \textcite{VanDenKerchove2024} and will be presented in
\cref{sec:covert-align}.
Performance is also evaluated on a publicly available dataset.

>>>>

The interface can be operated in different \ac{vsa} conditions, as illustrated in
\cref{fig:gaze/vsa}.
\begin{figure}
  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_overt.pdf}
    \caption[Overt \ac{vsa}]{%
      \emph{Overt} \ac{vsa}.
      Gaze and \ac{vsa} coincide on a target.
    }
    \label{fig:gaze/vsa/overt}
  \end{subfigure}\hfill%
  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_covert.pdf}
    \caption[Covert \ac{vsa}]{%
      \emph{Covert} \ac{vsa}.
      Gaze rests on the center of the screen, while \ac{vsa} is directed towards a target.
    }
    \label{fig:gaze/vsa/covert}
  \end{subfigure}

  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.pdf}
    \caption[Split \ac{vsa}]{%
      \emph{Split} \ac{vsa}.
      \Ac{vsa} is directed towards a target, while the gaze rests on another.
    }
    \label{fig:gaze/vsa/split}
  \end{subfigure}\hfill%
   \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.pdf}
    \caption[Free \ac{vsa}]{%
      \emph{Free} \ac{vsa}.
      The user is free to direct their gaze as they deem most comfortable.
      All previous \ac{vsa} conditions are possible.
    }
    \label{fig:gaze/vsa/free}
  \end{subfigure}\hfill%
  \caption[\Ac{vsa} conditions]{%
    \Acf{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm
    interface. Full lines indicated gaze fixation, dashed lines \ac{vsa}.
  }
  \label{fig:gaze/vsa}
\end{figure}
In the overt case, users gazed at the cued
target they were also mentally attending; in the covert case, users
gazed at the center of the screen while mentally attending to the cued target.
We introduced a \ac{vsa} condition that is understudied in the context of
gaze-independent \ac{bci} development: split \ac{vsa}.
In split \ac{vsa}, the user mentally focuses on one cued target while gazing at
another (the distractor).
This last option has been scarcely studied but completes
the options to dissociate gaze and visuospatial attention, allowing us to
investigate the effect of (the lack of) gaze control on \ac{bci} performance.
This dataset and the performance of our proposed \ac{wcble} decoder have also been
published by \textcite{VanDenKerchove2024} and will be presented in
\cref{sec:covert-align}.
Performance is also evaluated on a publicly available dataset.

### 2024-11-27 figures/bci/paradigms.tikz.tex
Bleed

    %% Text in rectangles
    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {\center+1.5*\textspacing}) {motor\\imagery};
    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {\center}) {imagined\\speech};
    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {\center-1.5*\textspacing}) {neurofeedback};

    \node[align=center,font=\footnotesize, color=muteblack] at ({\center+\textspacing}, {\center+\textspacing}) {P300};
    \node[align=center,font=\footnotesize, color=muteblack] at ({\center+\textspacing}, {\center-\textspacing}) {SSVEP};
    \node[align=center,font=\footnotesize, color=muteblack] at ({\center-\textspacing}, {\center+\textspacing}) {cVEP};
    \node[align=center,font=\footnotesize, color=muteblack] at ({\center-\textspacing}, {\center-\textspacing}) {mVEP};

    %% More text
    \node[align=center,font=\footnotesize, color=muteblack] at ({\center}, {-\center+\textspacing}) {error\\potentials};
    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {-\center+\textspacing}) {attention and \\ workload detection};
    \node[align=center,font=\footnotesize, color=muteblack] at ({-\center}, {-\center-\textspacing}) {clinical neuroimaging \\ and monitoring};
    \node[align=center,font=\footnotesize, color=muteblack] at ({+\center}, {-\center-\textspacing})  {emotion \\ detection};
  %\end{scope}
\end{tikzpicture}

>>>>

    %% Text in rectangles
    \node[align=center,font=\small, color=muteblack] at ({-\center}, {\center+1.5*\textspacing}) {motor\\imagery};
    \node[align=center,font=\small, color=muteblack] at ({-\center}, {\center}) {imagined\\speech};
    \node[align=center,font=\small, color=muteblack] at ({-\center}, {\center-1.5*\textspacing}) {neurofeedback};

    \node[align=center,font=\small, color=muteblack] at ({\center+\textspacing}, {\center+\textspacing}) {P300};
    \node[align=center,font=\small, color=muteblack] at ({\center+\textspacing}, {\center-\textspacing}) {SSVEP};
    \node[align=center,font=\small, color=muteblack] at ({\center-\textspacing}, {\center+\textspacing}) {cVEP};
    \node[align=center,font=\small, color=muteblack] at ({\center-\textspacing}, {\center-\textspacing}) {mVEP};

    %% More text
    \node[align=center,font=\small, color=muteblack] at ({\center}, {-\center+\textspacing}) {error\\potentials};
    \node[align=center,font=\small, color=muteblack] at ({-\center}, {-\center+\textspacing}) {attention and \\ workload detection};
    \node[align=center,font=\small, color=muteblack] at ({-\center}, {-\center-\textspacing}) {clinical neuroimaging \\ and monitoring};
    \node[align=center,font=\small, color=muteblack] at ({+\center}, {-\center-\textspacing})  {emotion \\ detection};
  %\end{scope}
\end{tikzpicture}
